A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-10-27 jour/jour.tex
more

Datatypes and codatatypes are an integral part of many proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, datatypes
have emerged in a few automatic theorem provers as well. The SMT-LIB
%\cite{barrett-et-al-2010}
format, implemented by most SMT
%(satisfiability modulo theories)
solvers, has been extended with a syntax for datatypes.
In this article, we introduce a
unified decision procedure for universal problems involving datatypes and codatatypes
in combination (Section~\ref{sec:the-decision-procedure}).
The procedure is described abstractly as a %tableau-like
calculus and is composable via the Nelson--Oppen method \cite{nelson-oppen-1979}.
It generalizes the procedure by Barrett et al.\ \cite{barrett-et-al-2007},
which covers only datatypes.
%To our knowledge, our procedure is the first of its kind for the theory of
%codatatypes.

@@

\paragraph{Conventions.}
Our setting is a monomorphic (or many-sorted) first-order logic.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$. Types are atomic sorts and interpreted as
nonempty domains. The set~$\Types$ must contain a
distinguished type \ty{bool} interpreted as the set of truth
values. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
The metavariables $\delta,\:\varepsilon$ range over (co)datatypes,
whereas $\tau,\:\upsilon$ range over arbitrary types.

@@

\newcommand\gandl{G\&L}
\newcommand\HD[1]{\hbox to2.25em{\footnotesize\hfill\!\!\!\!#1\!\!\!\!\hfill}}

@@

\newcommand\captioni{\,Number of solved goals for the three benchmark suites}
%\renewcommand\captioni{\,Number of solved goals with 16~lemmas per goal}

\begin{table*}[t!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}cc@{\kern.5em}cc@{\kern.5em}cc@{\kern1.3em}cc@{\;}}
  & \multicolumn{2}{c}{Distro} & \multicolumn{2}{c}{AFP} & \multicolumn{2}{c}{\gandl} & \multicolumn{2}{c}{Overall} \\[-0.5pt]
  & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4}& \HD{Z3} 
\MIDRULE
No (co)datatypes
  & 221 & 209 & 775 & 777 & 52 & 51 & 1048 & 1037 \\%[\jot]
%Datatype simplification only
%  &  \\
Datatypes without \rn{Acyclic}
  & 227 & -- & 780 & -- & 52 & -- & 1059 & -- \\
Full datatypes
  & 227 & 213 & 786 & 791 & 52 & 51 & 1065 & 1055 \\%[\jot]
%Codatatype simplification only
%  &  \\
Codatatypes without \rn{Unique}\!\!
  & 222 & -- & 804 & -- & 56 & -- & 1082 & -- \\
Full codatatypes
  & 223 & -- & 804 & -- & \bfseries 59 & -- & 1086 & -- \\%[\jot]
Full (co)datatypes
  & \bfseries 229 & -- & \bfseries 815 & -- & \bfseries 59 & -- & \bfseries 1103 & -- %\\[\jot]
\end{tabular}\end{center}
\caption{\captioni}
\label{tab:benchsixteen}
\end{table*}

>>>>

Datatypes and codatatypes are an integral part of many proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, datatypes
have emerged in a few automatic theorem provers as well. The SMT-LIB
\cite{xxxxxxx}
format, implemented by most SMT
%(satisfiability modulo theories)
solvers, has been extended with a syntax for datatypes.
In this article, we introduce a
unified decision procedure for universal problems involving datatypes and codatatypes
in combination (Section~\ref{sec:the-decision-procedure}).
The procedure is described abstractly as a %tableau-like
calculus and is composable via the Nelson--Oppen method \cite{nelson-oppen-1979}.
It generalizes the procedure by Barrett et al.\ \cite{barrett-et-al-2007},
which covers only datatypes.
%To our knowledge, our procedure is the first of its kind for the theory of
%codatatypes.

@@

\paragraph{Conventions.}
Our setting is a monomorphic (or many-sorted) first-order logic.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$\negvthinspace. Types are atomic sorts and interpreted as
nonempty domains. The set~$\Types$ must contain a
distinguished type \ty{bool} interpreted as the set of truth
values. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
The metavariables $\delta,\:\varepsilon$ range over (co)datatypes,
whereas $\tau,\:\upsilon$ range over arbitrary types.

@@

\newcommand\gandl{G\&L}
\newcommand\HD[1]{\hbox to2em{\footnotesize\hfill\!\!\!\!#1\!\!\!\!\hfill}}

@@

\newcommand\captioni{\,Number of solved goals with 16~lemmas per goal}
\newcommand\captionii{\,Number of solved goals with 256~lemmas per goal}

\begin{table*}[t!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1.5em}c@{\kern.5em}c@{\;}}
  & \multicolumn{2}{@{}c@{\kern1em}}{Distro} & \multicolumn{2}{@{}c@{\kern1em}}{AFP} & \multicolumn{2}{@{}c@{\kern1.5em}}{\gandl} & \multicolumn{2}{@{}c@{\;}}{Overall} \\[-0.5pt]
  & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4}& \HD{Z3} 
\MIDRULE
No (co)datatypes
  & 221 & 209 & 775 & 777 & 52 & 51 & 1048 & 1037 \\%[\jot]
%Datatype simplification only
%  &  \\
Datatypes without \rn{Acyclic}
  & 227 & -- & 780 & -- & 52 & -- & 1059 & -- \\
Full datatypes
  & 227 & 213 & 786 & 791 & 52 & 51 & 1065 & 1055 \\%[\jot]
%Codatatype simplification only
%  &  \\
Codatatypes without \rn{Unique}\!\!
  & 222 & -- & 804 & -- & 56 & -- & 1082 & -- \\
Full codatatypes
  & 223 & -- & 804 & -- & \bfseries 59 & -- & 1086 & -- \\%[\jot]
Full (co)datatypes
  & \bfseries 229 & -- & \bfseries 815 & -- & \bfseries 59 & -- & \bfseries 1103 & -- %\\[\jot]
\end{tabular}\end{center}
\caption{\captioni}
\label{tab:benchsixteen}
\end{table*}

\begin{table*}[t!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1.5em}c@{\kern.5em}c@{\;}}
  & \multicolumn{2}{@{}c@{\kern1em}}{Distro} & \multicolumn{2}{@{}c@{\kern1em}}{AFP} & \multicolumn{2}{@{}c@{\kern1.5em}}{\gandl} & \multicolumn{2}{@{}c@{\;}}{Overall} \\[-0.5pt]
  & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4}& \HD{Z3} 
\MIDRULE
No (co)datatypes
  & 221 & 209 & 775 & 777 & 52 & 51 & 1048 & 1037 \\%[\jot]
%Datatype simplification only
%  &  \\
Datatypes without \rn{Acyclic}
  & 227 & -- & 780 & -- & 52 & -- & 1059 & -- \\
Full datatypes
  & 227 & 213 & 786 & 791 & 52 & 51 & 1065 & 1055 \\%[\jot]
%Codatatype simplification only
%  &  \\
Codatatypes without \rn{Unique}\!\!
  & 222 & -- & 804 & -- & 56 & -- & 1082 & -- \\
Full codatatypes
  & 223 & -- & 804 & -- & \bfseries 59 & -- & 1086 & -- \\%[\jot]
Full (co)datatypes
  & \bfseries 229 & -- & \bfseries 815 & -- & \bfseries 59 & -- & \bfseries 1103 & -- %\\[\jot]
\end{tabular}\end{center}
\caption{\captionii}
\label{tab:benchtwofittysix}
\end{table*}

\newpage
